---
title: "The SiMBA study - Complete Analysis"
author: "Kaare D. Tranæs"
date: "5/08/2022"
output:
  html_document: default
  pdf_document: default
---

# The SiMBA study

### RCT study with 2 groups; FCS (active treatment) and Placebo. Measurements are from baseline and end of trial. All participants completede (no drop outs).

<!-- #ToDO list: -->

<!-- #Find out how to convert NA's to group means for line ( ) -->

<!-- #Change how variables are renamed. Use rename_vars() (or similar) and write variable names directly. -->

<!-- #Find out how to reference code inline -->

<!-- # -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_knit$set(root.dir = "../")
```

```{r load packages, include=FALSE}
library(readxl)
library(tidyverse)
library(visdat)
library(ruler)
library(knitr)
library(outliers)
library(tableone)
library(broom)
library(gt)
library(writexl)
library(patchwork)
```

## Biochemical analysis of clinical variables

#### Data import and cleaning

The data set 'simba working data 2' is imported and cleaned. \> Missing values in the raw data (empty cells) are stored in the variable MissingData, and are replaced with the mean value of the variable for each group.

The plot below confirmed that no data are missing going moving forward.

```{r load and clean data, echo=FALSE, warning=FALSE}

data  <- read_excel("RawData/simba working data 2.xlsx") %>% # Load data
  pivot_longer(                                   # Pivot to long format
               cols = weight_baseline:asat_end,
               names_to = "Variables",
               values_to = "Values") %>% 
  mutate(                                      # Change visit names
    Visit = ifelse(str_detect(Variables,'baseline'),'Baseline', 
            ifelse(str_detect(Variables,'end'),'End','miss')),
    Variables = Variables %>% str_remove('_baseline') %>% str_remove('_end') %>% 
           trimws() %>% tolower()) %>% 
  rename(                                      # Rename 
    CompletionStatus = status_end) %>% 
  mutate(
    Treatment = ifelse(str_detect(Treatment,'A'),'FCS',
                       ifelse(str_detect(Treatment,'B'),'Placebo','miss')),
         Treatment = as.factor(Treatment)) %>% 
  select(
    -status_baseline) %>% 
  relocate(
    Birthday, Age,Sex, CompletionStatus, Treatment, Visit, Variables,Values, .after=`Participant_ID`) %>% 
  group_by(                                     # Replace missing values with group mean
    Variables) %>% 
  mutate(
    Values = replace_na(Values, mean(Values, na.rm = TRUE))
  ) 

data %>%      # Check for missing values
  vis_miss()

```

```{r change variable names, eval=FALSE, include=FALSE}
data <- data %>% # These lines rename variables
  mutate(
    NewVariables = ifelse(str_detect(Variables,'peptidt0'), 'C-peptide 0m (pmol/L)',       
                  ifelse(str_detect(Variables,'peptidt30'), 'C-peptide 30m (pmol/L)',
                  ifelse(str_detect(Variables,'peptidt120'), 'C-peptide 120m (pmol/L)',
                  ifelse(str_detect(Variables,'glucoset0'), 'Glucose 0m (mmol/L)',
                  ifelse(str_detect(Variables,'glucoset30'), 'Glucose 30m (mmol/L)',
                  ifelse(str_detect(Variables,'glucoset120'), 'Glucose 120m (mmol/L)',
                  ifelse(str_detect(Variables, 'insulin_t0'), 'Insulin 0m (pmol/L)',
                  ifelse(str_detect(Variables, 'insulin_t30'), 'Insulin 30m (pmol/L)',
                  ifelse(str_detect(Variables, 'insulin_t120'), 'Insulin 120m (pmol/L)',
                  ifelse(str_detect(`Variables`,'weight'),'Weight (kg)',
                  ifelse(str_detect(`Variables`,'fatmass'),'Fatmass (kg)',
                  ifelse(str_detect(`Variables`,'waist'),'Waist (cm)',
                  ifelse(str_detect(`Variables`,'height'),'Height (cm)',
                  ifelse(str_detect(`Variables`,'bmi'),'BMI (kg/m2)',
                  ifelse(str_detect(`Variables`,'trig'),'Triglycerides (mmol/l)',
                  ifelse(str_detect(`Variables`,'ldl'),'LDL-Cholesterol (mmol/l)',
                  ifelse(str_detect(`Variables`,'hdl'),'HDL-Cholesterol (mmol/l)',
                  ifelse(str_detect(`Variables`,'chol'),'Total Cholesterol (mmol/l)',
                  ifelse(str_detect(`Variables`,'hba1c_h'),'HbA1c Hemoglobin',
                  ifelse(str_detect(`Variables`,'hba1c_h'),'HbA1c Hemoglobin',
                  ifelse(str_detect(`Variables`,'hba1c_hemolysat'),'HbA1c (DCCT%)',
                  ifelse(str_detect(`Variables`,'crp'),'C-reactive Protein (mg/L)',
                  ifelse(str_detect(`Variables`,'alat'),'ALAT (U/L)',
                  ifelse(str_detect(`Variables`,'asat'),'ASAT (U/L)',
                  ifelse(str_detect(`Variables`,'auc'),'Glucose AUC',
                  Variables)))))))))))))))))))))))))) %>%
   mutate(NewVariables = factor(NewVariables, levels = c("Weight (kg)","Height (cm)","BMI (kg/m2)","Waist (cm)","Fatmass (kg)", "C-reactive Protein (mg/L)", "Triglycerides (mmol/l)", "LDL-Cholesterol (mmol/l)", "HDL-Cholesterol (mmol/l)", "Total Cholesterol (mmol/l)", "Glucose 0m (mmol/L)", "Glucose 30m (mmol/L)", "Glucose 120m (mmol/L)", "HbA1c (DCCT%)", "Insulin 0m (pmol/L)", "Insulin 30m (pmol/L)", "Insulin 120m (pmol/L)", "C-peptide 0m (pmol/L)", "C-peptide 30m (pmol/L)", "C-peptide 120m (pmol/L)", "ALAT (U/L)", "ASAT (U/L)", "Glucose AUC"))) %>%
  arrange(NewVariables)

```

### Outlier inspection

The code below calculates a z-score for all observations and determines values with z-scores \>4 as outliers. All varialbes are then plotted as histograms for visual inspection.

Although there are some statistical outliers, they are not biologically implausible and are thus kept in the data analysis.

```{r Check for outliers, echo=FALSE}

outliers <- function(x, thres = 4, na.rm = TRUE) {
  s <- sd(x, na.rm = na.rm) 
  mn <- mean(x, na.rm = na.rm)
  o <- rep(FALSE,length(x))
  o[x < (mn - thres*s)] <- TRUE
  o[x > (mn + thres*s)] <- TRUE
  o[is.na(x)] <- TRUE
  return(o)
}

data %>% 
  group_by(Variables) %>% 
  mutate(e = lm(Values~ Visit*Treatment + Sex) %>% residuals(),
         Outlier = outliers(e))%>% 
  ggplot(aes(Values, fill = Outlier)) + 
  geom_histogram()  + 
  facet_wrap(~Variables,scales =  'free')

#data <- data %>% 
#   filter(Outlier == FALSE)
```

### Table One

The following table shows the two intervention groups at baseline. Purpose is to show that randomization worked, and hence that the two groups are not different.

```{r Table One, echo=FALSE}

#baseline <-            ####### Write baseline stats to csv if needed by uncommenting this line ###### 
data %>% 
  filter(Visit == "Baseline") %>% 
    pivot_wider(names_from = "Variables", values_from = "Values") %>% 
 CreateTableOne(., strata = "Treatment", vars = c('Age','Sex', 'weight', 'height','fatmass','waist','trig','hdl','ldl','chol','glucoset0','glucoset30','glucoset120','auc','crp')) %>% 
print(., nonnormal = 'Age', cramVars = 'Sex')
 # %>% baseline_csv <- print(baseline, nonnormal = 'Age', cramVars = 'Sex') ####### Write baseline stats to csv if needed by uncommenting this line ###### 
# write.csv(baseline_csv, file = "../RawData/SimbaBaseline.csv")
```

The table above illustrates that the groups were evenly matched at baseline.

Baseline values of interest are illustrated below.

```{r baseline values displayed}
data %>% 
  filter(Visit == 'Baseline') %>% 
  filter(Variables %in% c('weight','glucoset0','glucoset30','glucoset120','crp', 'fatmass')) %>% 
  ggplot(data = ., aes(x = Treatment,
                       y = Values,
                       color = Treatment,
                       #group = factor(Participant_ID)
                       )) + 
  geom_jitter() + theme(legend.position = NULL) + facet_wrap(~Variables, scales = 'free')
```

### Results tabel

Below is the results of a linear regression model calculating the effect of the intervention on changes from baseline to end of trial. Results are displayed as Placebo VS Treatment (so a negative value means an increase in the treatment group).

```{r linear model, echo=FALSE}
# data %>%                                                         Create excel-table with summary statistics
#   select(-Birthday,-CompletionStatus) %>% 
#   group_by(Treatment, Visit, Variables) %>% 
#   summarise(mean = mean(Values), 
#             sd = sd(Values)) %>% 
#   write_xlsx(, path = "Simba summary statistics all sexes w. p-value.xlsx")

data %>% 
  select(-CompletionStatus) %>% 
  group_by(Participant_ID, Treatment, Sex, Variables) %>% 
  summarise(Change = Values[Visit== "End"]- Values[Visit== "Baseline"]) %>% 
  mutate(All = "All") %>% 
  gather(sx,Sex,All,Sex) %>% 
  group_by(Variables,Sex) %>% 
  do(lm(data = ., Change ~ Treatment) %>%
       tidy) %>% 
  filter(term!='(Intercept)') %>% 
  select(-statistic, -term) %>% 
  kable(., caption = 'Placebo versus Treatment. Effect of intervention reported for both genders and seperately)', digits = 4)
```

### Primary study outcomes visualized

> Blood glucose levels for each treatment group during 2-hour OGTT

```{r primary outcome plot}
visit.labs <- c("Pre-trial", "Post-trial") # Names you want on plot
names(visit.labs) <- c("Baseline", "End") # Current names of visits in data

################## Boxplot showing 0, 30 and 120 min glucsoe levels. Groups side-by-side. Faceted by visit. ###########
all <- data %>% 
  filter(Variables == "glucoset0" | Variables == "glucoset30"| Variables == "glucoset120") %>% 
  ggplot(., aes(x = factor(Variables, level= c("glucoset0", "glucoset30", "glucoset120")),   # Change order of variables
                y = Values,
                color = Treatment)) + 
  geom_boxplot() + facet_wrap(~Visit,
               labeller = labeller(Visit = visit.labs),
             ncol = 2) +
  scale_x_discrete(labels=c("glucoset0" = "0 min",
                             "glucoset30" = "30 min",
                             "glucoset120"= "120 min"
                             ))+
   ylab("Glucose, mmol/L")+ 
   xlab("OGTT time points") +
  ggtitle("Oral Glucose Tolerance Test") 

################## Line plot showing time course of glucose during OGTT. Pre and Post in same plot. Faceted by Group. ###########

OGTT <- data %>% 
  filter(Variables == "glucoset0" | Variables == "glucoset30"| Variables == "glucoset120") %>% 
  group_by(Treatment, Visit, Variables) %>%  # Calculating mean for both groups 
  summarise(MeanValues = mean(Values)) %>%   # Calculating mean for both groups
  ggplot(., aes(x = factor(Variables, level= c("glucoset0", "glucoset30", "glucoset120")),  # Change order of variables
                y = MeanValues,
                group = Visit,
                color = Visit)) + 
  geom_point() + geom_line() + theme(legend.position = NULL) + facet_wrap(~Treatment,
             ncol = 2) +
  scale_color_discrete(labels=c('Pre-trial', 'Post-trial')) +
  scale_x_discrete(labels=c("glucoset0" = "0 min",
                             "glucoset30" = "30 min",
                             "glucoset120"= "120 min"
                             ))+
   ylab("Glucose, mmol/L")+ 
   xlab("OGTT time points")

################## Boxplot showing AUC at Pre and Post. Faceted by Group. ###########

AUC <- data %>% 
  filter(Variables == "auc") %>% 
  ggplot(., aes(x = Visit, 
                y = Values,
                #grouping = Variables,
                fill = Treatment)) + 
  geom_boxplot() + facet_wrap(~Treatment,
             ncol = 2) +
  scale_x_discrete(labels=c("Baseline" = "Pre-trial", "End" = "Post-trail"))+
   ylab("Glucose, mmol/L")+ 
   xlab("Visits")

##### Patchwork ###### 

patchwork <- OGTT / AUC
PatchOut <- patchwork + plot_annotation(
  title = "Blood Glucose Concentration Levels at Pre-trial and Post-trial",
  tag_levels = 'A' 
)
#ggsave('OGTT results.pdf', PatchOut, limitsize = FALSE, height = 10, width = 13)
```


```{r changes displayed, eval=FALSE, include=FALSE}
data %>% 
  #filter(time %in% c('baseline','end')) %>% 
  filter(Variables %in% c('age','weight','glucoset0','glucoset30','glucoset120','crp', 'fatmass')) %>% 
  ggplot(data = ., aes(x = Visit,
                       y = Values,
                       color = Treatment,
                       group = factor(Participant_ID)
                       )) + 
  geom_point() + geom_line() + theme(legend.position = NULL) + facet_wrap(~Variables, scales = 'free')
```

## Metabolomics of Plasma Fatty Acids and Amino Acids

### Import Fatty Acid Raw Data

```{r load fame data}
fame <- read_excel("RawData/FAME_working_copy.xlsx", skip = 2) %>% 
  mutate(Visit = ifelse(str_detect(`Sample ID`,'_1'),'Baseline',
                       ifelse(str_detect(`Sample ID`,'_2'),'End','miss'))) %>% mutate('Sample ID' = `Sample ID` %>% str_remove('_1') %>% str_remove('_2'))
  
```

#### Load SIMBA ID coupling list

```{r load coupling list}

treatment_list <- read_excel("RawData/SIMBA_ID_KOBLING.xlsx",
                      col_types = c("numeric", "text", "text"))

participant_info <- read_excel("RawData/Participant_info.xlsx",
                      col_types = c("text", "numeric", "text", "text", "text"))

id <- inner_join(treatment_list, participant_info, by = "allocation_no") %>% select(-Id, -Lastname)

```

### Join datasets

```{r assign intervention}
#Join fame-data with participant allocation list                                        #Problem with allocation_no. No ideas. FIxed! 
fame <- fame %>% 
  dplyr::inner_join(., id, by = c('Sample ID' = 'allocation_no')) %>% 
  mutate(treatment = ifelse(str_detect(treatment,'A'),'FCS',
                       ifelse(str_detect(treatment,'B'),'Placebo','miss'))) %>% 
  relocate(Visit, .after=`Sample ID`) %>% 
  relocate(treatment, .after=`Sample ID`)%>% 
  relocate(Age:Sex, .after=`Sample ID`) %>% 
  select(-Randomiseringsid)
head(fame)
```

#### Wrap to long format

```{r, wrap to long format}
fame_long <- fame_join %>% 
  pivot_longer("C10:0":"C24:1", 
               names_to = "Fatty_Acids", 
               values_to = "µmol/L")
head(fame_long)
```

<!-- ## Corona sub-analysis -->

<!-- ```{r eval=FALSE, include=FALSE} -->

<!-- #corona <- read_excel("files to bind/Corona-affected participants.xlsx") -->

<!-- #corona_participants <- corona %>%  -->

<!--  # left_join(biomarkers, by = c("Corona_participants" = "Participant")) -->

<!-- ``` -->

<!-- # Effect of treatment of corona affected subjects - linear regression -->

<!-- ```{r} -->

<!-- # tb_corona <- corona_participants %>%  -->

<!-- #   #filter(Participant == corona$Corona_participants) %>%  -->

<!-- #   group_by(Variables, Corona_participants, Treatment,Sex) %>%  -->

<!-- #   summarise(Change = Values[Visit=='End'] - Values[Visit=='Baseline']) %>%  -->

<!-- #   mutate(All = 'All') %>%  -->

<!-- #   pivot_longer(cols = c(All, Sex), names_to = "sex", values_to = "Sex") %>%  -->

<!-- #   group_by(Variables,Sex) %>%  -->

<!-- #   do(lm(data = ., Change ~ Treatment) %>% tidy)  %>% filter(term!='(Intercept)') -->

<!-- #       #### effect estimate for lm(change) = -1 x effect estimate -->

<!-- #  -->

<!-- # tb_corona %>% select(-term, -statistic) %>%  -->

<!-- #   kable(., caption = 'change from baseline results treatmentB versus treatmeentA (reported for all and both genders seperately)', digits = 4) -->

<!-- #  -->

<!-- # tb_corona %>% rio::export(.,file = "_corona_affected_results_change_both_sexes.xlsx") -->

<!-- ``` -->

<!-- # Effect of treatment - linear regression -->

<!-- ```{r} -->

<!-- # tb <- biomarkers %>%  -->

<!-- #   #filter(Participant == corona$Corona_participants) %>%  -->

<!-- #   group_by(Variables, Participant,Treatment,Sex) %>%  -->

<!-- #   summarise(Change = Values[Visit=='End'] - Values[Visit=='Baseline']) %>%  -->

<!-- #   mutate(All = 'All') %>%  -->

<!-- #   pivot_longer(cols = c(All, Sex), names_to = "sex", values_to = "Sex") %>%  -->

<!-- #   group_by(Variables,Sex) %>%  -->

<!-- #   do(lm(data = ., Change ~ Treatment) %>% tidy)  %>% filter(term!='(Intercept)') -->

<!-- #       #### effect estimate for lm(change) = -1 x effect estimate -->

<!-- #  -->

<!-- # tb %>% select(-term, -statistic) %>%  -->

<!-- #   kable(., caption = 'change from baseline results treatmentB versus treatmeentA (reported for all and both genders seperately)', digits = 4) -->

<!-- #  -->

<!-- # #tb %>% rio::export(.,file = "results_change_both_sexes.xlsx") -->

<!-- ``` -->

<!-- #### Plot results -->

<!-- ```{r plot results and inspect for outliers} -->

<!-- # #all_plots <- biomarkers %>%  -->

<!-- #   filter(Visit != "Change") %>%  -->

<!-- #   ggplot(., aes(x = Variables,  -->

<!-- #                 y = Values,  -->

<!-- #                 fill = Treatment)) +  -->

<!-- #   geom_boxplot() + -->

<!-- #   facet_wrap(~Variables + Visit,  -->

<!-- #              scales = "free",  -->

<!-- #              ncol = 2) -->

<!-- # #all_plots -->

<!-- # ggsave('Inspect plots box.pdf', all_plots, limitsize = FALSE, height = 70) #Open file and inspect.  -->

<!-- ``` -->

<!-- #### See significant results -->

<!-- ```{r, calculate significant stats} -->

<!-- # significant_results <- tb %>%  -->

<!-- #   select(-statistic, -term) %>%  -->

<!-- #   filter(p.value <= 0.05) -->

<!-- # significant_results -->

<!-- ``` -->

<!-- ### Plot Significant results -->

<!-- ```{r} -->

<!-- # biomarkers %>%  -->

<!-- #   filter(Variables == significant_results$Variables) %>%  -->

<!-- #   filter(!Visit %in% "Change") %>%  -->

<!-- #   ggplot(., aes(x = Visit,  -->

<!-- #                 y = Values, -->

<!-- #                 grouping = Treatment, -->

<!-- #                 color = Sex)) +  -->

<!-- #   geom_boxplot() +  -->

<!-- #   ggtitle("Fasting Blood Glucose") + -->

<!-- #   xlab("Visit") + -->

<!-- #   ylab("mmol/L") -->

<!-- ``` -->

<!-- ```{r save results, include=FALSE, echo=FALSE} -->

<!-- #ggsave('Significant Plots.pdf', significant_plots, limitsize = FALSE, height = 8) -->

<!-- ``` -->
